\Title

### 2023-02-10 Grimaldi-etal-BiolCybernetics.tex
Update on Overleaf.
Update Grimaldi-etal-BiolCybernetics.tex
methods

In neurobiological recordings or in the sensory signal obtained from an event-based camera, any generic raster plot consists of a stream of \emph{spikes} or \emph{events}. This can be formalized as a list of neural addresses and timestamps tuples $\event = \{(\presynaddr_\arank, \timev_\arank)\}_{\arank \in [1,\numevent]}$ where $\numevent \in \mathbb{N} $ is the total number of events in the data stream and the rank $\arank$ is the index of each event in the list of events. Events are typically ordered by their time of occurrence. Each event has a time of occurrence $\timev_\arank$ and an associated address $\presynaddr_\arank$. This defines an address space $\presynaddrspace$ which consists of the set of possible addresses. In a neurobiological recording, this can be the identified set of neurons. For event-based cameras, this can be defined as $[1, \Npol] \times [1, \Nx]\times[1, \Ny] \subset \mathbb{N}^3$ where $\Npol$ is the number of polarities ($\Npol=2$ for the ON and OFF polarities coded in event-based cameras) and $(\Nx, \Ny)$ is the size of the sensor in pixels. As such, an address $\presynaddr_\arank$ is typically in the form $(\polev_\arank, x_\arank, y_\arank)$ for event-based cameras.

Let's formalize the heterogeneous delays spiking neural network (HD-SNN). Each neuron $\postsynaddr \in \postsynaddrspace$ connects to presynaptic afferents from $\presynaddrspace$. In biology, a single cortical neuron has generally several thousands of synapses. Each synapse may be defined by its synaptic weight and its delay, that is, the time it takes for one spike to travel from the presynaptic neuron's soma to that of the postsynaptic neuron. A postsynaptic neuron $\postsynaddr \in \postsynaddrspace$ is then described by the synaptic weights connecting it to a presynaptic afferent from $\presynaddrspace$ but also by the set of possible delays. Note that a neuron may contact an afferent neuron with multiple different delays. Scanning all neurons $\postsynaddr$, we thus define the full set of $\Nsyn$ synapses, 
as  $\synapse = \{(\presynaddr_\ranksyn, \postsynaddr_\ranksyn, \synapticweight_\ranksyn, \synapticdelay_\ranksyn)\}_{\ranksyn \in [1,\Nsyn]}$, where each synapse is associated to a presynaptic address $\presynaddr_\ranksyn$, a postsynaptic address $\postsynaddr_\ranksyn$, a weight $\synapticweight_\ranksyn$, and a delay $\synapticdelay_\ranksyn$. This defines the full connectivity of the HD-SNN model.

@@

\begin{equation}\label{eq:stream_b}
\event_\postsynaddr = \{(\presynaddr_\arank, \synapticweight_\arank, \timev_\arank+\synapticdelay_\ranksyn) \| \presynaddr_\arank = \presynaddr_\ranksyn \}_{\arank \in [1,\numevent], \ranksyn \in \synapse^\postsynaddr}
\end{equation}
% Unlike the binary stream of events from the signal as input, the postsynaptic neurons receive a weighted stream of event such that for each event $\arank$ its value is $\synapticweight_\arank = \synapticweight_\ranksyn$ when $\presynaddr_\arank = \presynaddr_\ranksyn$.
This new stream of events is ordered in time in biology as events reach the soma of post-synaptic neurons but should be reordered in simulations. Crucially, when post-synaptic neurons are activated on their soma by a specific spatio-temporal motif which is imprinted in the set of synapses, the discharge probability will increase, notably when these spikes converge on the soma in a synchronous manner while the activation function of the neurons of the HD-SNN can be selected among the whole range of spiking neuron response functions (see Figure~\ref{fig:izhikevich}). We defined the HD-SNN model in this subsection in all generality and in the next subsection we describe an implementation of our model adapted for the motion detection task. 
% \note{I can make a small figure with event streams of different colors for the input one, the multiplexed one etc and the response in the bottom}
\subsection{Application of HD-SNN to the motion detection task}
From the perspective of simulating such event-based computations on standard CPU- or GPU-based computers and then using parallel computing, it is useful to transform this event-based representation into a dense representation. Indeed, by discretization of time, we may transform any event-based input as a boolean matrix $A \in \{0, 1 \}^{\Npol \times \Ntime \times \Nx \times \Ny}$, defined for all polarities $p$, times $t$ and space coordinates $x$ and $y$. The values are by definition equal to zero except at the occurrences of events: $\forall \numevent \in \mathbb{N}$, $A(\polev_\arank, t_\arank, x_\arank, y_\arank)=1$. First, one can note that by using a discretization, the computational brick used in equation~\eqref{eq:kernel_b} corresponds to a temporal convolution which transforms the input $A$ using one kernel per postsynaptic neuron~\citep{grimaldi_learning_2022}. To benefit from invariance to location observed in images and exploited in Convolutional Neural Networks, one may further assume that synaptic motifs should be similar across different positions such that we may define a spatio-temporal convolution operator. % The characteristics of the kernels are shared among one class of neurons such that $\forall i,j \in [1, N_\class]^2, \; \kernel^i_\class = \kernel^j_\class$. 
% Here, we will consider that heterogeneous delays are limited in range
In particular, the whole synaptic set can be represented as a one kernel for each class $\class$ of the supervision task as the dense matrix $\kernel_\class$ of size $(\Npol, \Ktime, \Kx, \Ky)$, where  $\Ktime$ is the number of delays and $\Kx$ and $\Ky$ are the number of pixels in both spatial dimensions. To keep an analogy with the HD-SNN model, for each neuron $\postsynaddr$ of position $(x, y)$ and class $\class$, $\kernel_\class$ gives the weights as a function of presynaptic address and delay: $\forall {\ranksyn \in \synapse^\postsynaddr}, \kernel_\class(\presynaddr_\ranksyn, \synapticdelay_\ranksyn) = \synapticweight_\ranksyn$. %Therefore, if we make the approximation that one neuron integrates only on the temporal window given by its variety of synaptic delays, the integration of the spike train as input can be formalized by a 3D convolution operation. The longest synaptic delay defines the depth of the kernel $\kernel^\postsynaddr$ and all possible delays associated to the different presynaptic addresses are represented. 
Using this dense representation, the processing of $A$ by the model can be written as:
\begin{equation}\label{eq:kernel_b}
\forall x, y, \timev, \; \;
\mathcal{C}_\class(x, y, \timev)
= \sum_{p, \delta_x, \delta_y, \synapticdelay} \kernel_\class(p, \delta_x, \delta_y, \synapticdelay) \cdot A(p, x - \delta_x, y - \delta_y, \timev-\synapticdelay)
\end{equation}
% The computational brick used in equation~\eqref{eq:kernel_b} can be extended to every position, in space and time, of the boolean matrix $A$: equation~\eqref{eq:stream_b}
% \begin{equation}\label{eq:conv}
% \forall \presynaddr \in \presynaddrspace, \forall \timev \in T, \; \;
% \mathcal{C}_\class(\presynaddr,\timev)
% = \sum_{\ranksyn  \in [1, N_{\synapse^\postsynaddr}]} \kernel_\class(\presynaddr_\ranksyn, \synapticdelay_\ranksyn) \cdot A(\presynaddr, \timev-\synapticdelay_\ranksyn)
% \end{equation}
This shows that $\mathcal{C}_\class$ is a spatio-temporal convolution of the dense representation of the event stream with the dense kernels formed by the set of synapses:  $\mathcal{C}_\class = \kernel_\class \ast A$ (see Figure~\ref{fig:model} for an illustration). Note that to remain within the framework of a causal calculation, the kernels are shifted in time such that only past information gives an answer at the present time.
This well-known computation defines a differentiable measure which is very efficiently implemented for GPUs and which we will use for learning the classification of different motifs in the event stream. The same type of spatio-temporal filtering is used as a preprocessing stage of a pattern recognition algorithm~\citep{ghosh_spatiotemporal_2019}. Also,~\citet{sekikawa_constant_2018} developed an efficient 3D convolutional algorithm which implements a motion estimation task. By assuming a locally constant velocity, the authors assume the 3D kernel can be decomposed into a 2D kernel representing the shapes and a 3D kernel representing the velocity.

Here, we keep the analogy with spiking neurons, and we try to use as few heuristics as possible to observe the emergence of spiking motifs in the spatio-temporal kernels. If so, this precise spatio-temporal patterns prove to be of interest for neural computations and one can assume that biological neurons make use of this information as well. 
% \note{plus de biblio sur la motion detection/segmentation ou optique flow avec les events? (reviewer 2)}
The 3D convolution represents the linear integration of the spike train as input. The activation function of our model is a softmax function implementing a form of Multinomial Logistic Regression (MLR)~\citep{grimaldi_robust_2022}, in analogy to a spiking Winner-Take-All network~\citep{nessler_bayesian_2013}. In our MLR model, a probability value, for each class (i.e. each motion direction), is predicted for each event at position $x, y$ and at time $\timev$ as a softmax function of the linear combination of the list of events. % on the basal dendrite of a neuron $\postsynaddr$ in association to a specific class. The linear combination can be defined by a set of synapses $\synapse^\postsynaddr$ as described in the HD-SNN model. 
Using the kernels, it transforms the input raster plot into a probability with the following formula:
\begin{equation}\label{eq:mlr}
    \forall x, y, \timev, \forall \class \in [1, N_\class], \; \;
Pr(k=\class \; \vert \;  x, y, \timev) =
\frac {\exp  (\mathcal{C}_\class(x, y, \timev) +\bias_\class) }{Z(x, y, \timev)}
\end{equation} 
where $Z(x, y, \timev)=\sum_{\class \in [1, N_\class]} \exp  (\mathcal{C}_\class(x, y, \timev) +\bias_\class)$ is the partition function and  $\bias_\class$ is the bias linked to class $\class$. 
In particular, we expect that some specific motifs may become tightly synchronized as they reach the basal dendritic tree, leading to a high postsynaptic activity which makes it progressively more likely to generate an output spike. The spiking output of the model corresponds to an event with the highest probability class.

Once explained the general framework used to solve the ecological task described in section~\ref{sec:task}, we can add some heuristics based on neuroscientific observations to constrain our model and its strategies to solve the task. Note that the general framework is the same as the one presented in~\cite{grimaldi_learning_2022} and that, appart from the task to solve and a deeper analysis of the results, these priors, inspired by neuroscience, are the main methodological differences with this previous study. We set \note{arbitrarily} the size of the kernels to $(9, 17, 17)$ and define as many classes as the number of motion direction: $N_\class = 8$. Because receptive fields of biological neurons are mainly circular and can be more or less elongated, we apply a circular mask on the spatial dimensions of the kernels. And because the synaptic delay is linked, by physical principles, to the distance between the presynaptic neuron and the postsynaptic one, we apply a linear mask on the temporal axis such that the connectivity depends on the delay: the lower the delay the lower the circular mask radius (see Figure~\ref{fig:kernels} for an illustration). The second heuristics is based on the observation that biological neurons have a sparse spiking activity. From this observation, we set an output firing rate for the layer of spiking neurons in order to make a sparse selection of the events at which the model makes a decision. It results in the selection of the coefficients with the maximum likelihood value. Therefore, we will only select a given proportion of the most active cells, which we set here to $.01\%$ of all output voxels. This threshold was found using cross-validation and future work should automate how this may be determined for various datasets. Given a stimulus input, the model thus yields a spike output in the postsynaptic address space.

@@

\begin{figure}%[h!]%[ht!]
    \centering
    \includegraphics[width=\linewidth]{figures/HD-SNN.png}
    \caption{
    \textbf{- HD-SNN illustration. From left to right:} A 3D representation of the event stream as output of the DVS. A 3D convolution is applied, on the dense representation of the input, along the $X$ and $Y$ spatial axes (defining the DVS pixel grid) and the temporal axis. There are as many kernels as the possible number of motion classes in the recording. For clarity, only $4$ kernels are illustrated and some weights are not represented in the voxel grid. For each voxel on the dense representation of the event stream as input, the output of the 3D convolution is processed by the non-linearity of the MLR model (i.e. softmax function). The output of the MLR gives a probability for each class, associated to a specific kernel, to emit an output event, or spike. With an appropriate spiking mechanism, here a winner-takes-all, we obtain, as output of the HD-SNN model, a new spike train with the different spikes associated to a specific motion class. }
    \label{fig:model}
\end{figure}

@@

Since the model is fully differentiable, we can now implement a semi-supervised learning rule. The loss function of the MLR model is the binary cross-entropy on the output of the classification layer. Supervision was implemented using the input binary events as defined above and the motion direction labels as the desired output. The labels were defined at each time point as a one-hot encoding of the current motion in the channel corresponding to the current motion for all positions. As in this semi-supervised context, the label is known, but the timing is not, we used a selection of the temporal support in an unsupervised manner. It is important to note that we weighted the cost function by considering only active cells, such that the error is only back propagated to the spatial locations of these most active cells. This is reminiscent of previous methods solving this problem using a Winner-Takes-All mechanism~\citep{masquelier_unsupervised_2007}. Simulations are performed with the PyTorch library using gradient descent with Adam (for $2^{12}$ movies and a learning rate of $10^{-5}$). We demonstrated in a previous work that this type of learning algorithm can be assimilated to a Hebbian learning mechanism~\cite{grimaldi_robust_2022}.  

>>>>

In neurobiological recordings or in the sensory signal obtained from an event-based camera, any generic raster plot consists of a stream of \emph{spikes} or \emph{events}. This can be formalized as a list of neural addresses and timestamps tuples $\event = \{(\presynaddr_\arank, \timev_\arank)\}_{\arank \in [1,\numevent]}$ where $\numevent \in \mathbb{N} $ is the total number of events in the data stream and the rank $\arank$ is the index of each event in the list of events. Events are typically ordered by their time of occurrence. Each event has a time of occurrence $\timev_\arank$ and an associated address $\presynaddr_\arank$. This defines an address space $\presynaddrspace$ which consists of the set of possible addresses. In a neurobiological recording, this can be the identified set of presynaptic neurons. For event-based cameras, this can be defined as $[1, \Npol] \times [1, \Nx]\times[1, \Ny] \subset \mathbb{N}^3$ where $\Npol$ is the number of polarities ($\Npol=2$ for the ON and OFF polarities coded in event-based cameras) and $(\Nx, \Ny)$ is the size of the sensor in pixels. As such, an address $\presynaddr_\arank$ is typically in the form $(\polev_\arank, x_\arank, y_\arank)$ for event-based cameras.

In the HD-SNN, each neuron $\postsynaddr \in \postsynaddrspace$ connects to presynaptic afferents from $\presynaddrspace$. In biology, a single cortical neuron has generally several thousands of synapses. Each synapse may be defined by its synaptic weight and its delay, that is, the time it takes for one spike to travel from the presynaptic neuron's soma to that of the postsynaptic neuron. A postsynaptic neuron $\postsynaddr \in \postsynaddrspace$ is then described by the synaptic weights connecting it to a presynaptic afferent from $\presynaddrspace$ but also by the set of possible delays. Note that a neuron may contact an afferent neuron with multiple different delays. Scanning all neurons $\postsynaddr$, we thus define the full set of $\Nsyn$ synapses, 
as  $\synapse = \{(\presynaddr_\ranksyn, \postsynaddr_\ranksyn, \synapticweight_\ranksyn, \synapticdelay_\ranksyn)\}_{\ranksyn \in [1,\Nsyn]}$, where each synapse is associated to a presynaptic address $\presynaddr_\ranksyn$, a postsynaptic address $\postsynaddr_\ranksyn$, a weight $\synapticweight_\ranksyn$, and a delay $\synapticdelay_\ranksyn$. This defines the full connectivity of the HD-SNN model.

@@

\begin{equation}\label{eq:stream_b}
\event_\postsynaddr = \{(\presynaddr_\arank, \synapticweight_\arank, \timev_\arank+\synapticdelay_\ranksyn) \| \presynaddr_\arank = \presynaddr_\ranksyn \}_{\arank \in [1,\numevent], \ranksyn \in \synapse^\postsynaddr}
\end{equation}

% Unlike the binary stream of events from the signal as input, the postsynaptic neurons receive a weighted stream of event such that for each event $\arank$ its value is $\synapticweight_\arank = \synapticweight_\ranksyn$ when $\presynaddr_\arank = \presynaddr_\ranksyn$.
This new stream of events is ordered in time in biology as events reach the soma of post-synaptic neurons but should be reordered in simulations. Crucially, when post-synaptic neurons are activated on their soma by a specific spatio-temporal motif which is imprinted in the set of synapses, the discharge probability will increase, notably when these spikes converge on the soma in a synchronous manner while the activation function of the neurons of the HD-SNN can be selected among the whole range of spiking neuron response functions (see Figure~\ref{fig:izhikevich}). We defined the HD-SNN model in this subsection in all generality and in the next subsection we describe an implementation of our model adapted for the motion detection task. 
%
% \note{I can make a small figure with event streams of different colors for the input one, the multiplexed one etc and the response in the bottom}
%
\subsection{Application of HD-SNN to the motion detection task}
%
From the perspective of simulating such event-based computations on standard CPU- or GPU-based computers and then using parallel computing, it is useful to transform this event-based representation into a dense representation. Indeed, by discretization of time, we may transform any event-based input as a boolean matrix $A \in \{0, 1 \}^{\Npol \times \Ntime \times \Nx \times \Ny}$, defined for all polarities $p$, times $t$ and space coordinates $x$ and $y$. The values are by definition equal to zero except at the occurrences of events: $\forall \arank \in [1,\numevent]$, $A(\polev_\arank, t_\arank, x_\arank, y_\arank)=1$. First, one can note that by using a discretization, the computational brick used in equation~\eqref{eq:kernel_b} corresponds to a temporal convolution which transforms the input $A$ using one kernel per postsynaptic neuron~\citep{grimaldi_learning_2022}. To benefit from invariance to location observed in images and exploited in Convolutional Neural Networks, one may further assume that synaptic motifs should be similar across different positions such that we may define a spatio-temporal convolution operator. % The characteristics of the kernels are shared among one class of neurons such that $\forall i,j \in [1, N_\class]^2, \; \kernel^i_\class = \kernel^j_\class$. 
% Here, we will consider that heterogeneous delays are limited in range
In particular, the whole synaptic set can be represented as a one kernel for each class $\class$ of the supervision task as the dense matrix $\kernel_\class$ of size $(\Npol, \Ktime, \Kx, \Ky)$, where  $\Ktime$ is the number of delays and $\Kx$ and $\Ky$ are the number of pixels in both spatial dimensions. To keep an analogy with the HD-SNN model, for each neuron $\postsynaddr$ of position $(x, y)$ and class $\class$, $\kernel_\class$ gives the weights as a function of presynaptic address and delay: $\forall {\ranksyn \in \synapse^\postsynaddr}, \kernel_\class(\presynaddr_\ranksyn, \synapticdelay_\ranksyn) = \synapticweight_\ranksyn$. %Therefore, if we make the approximation that one neuron integrates only on the temporal window given by its variety of synaptic delays, the integration of the spike train as input can be formalized by a 3D convolution operation. The longest synaptic delay defines the depth of the kernel $\kernel^\postsynaddr$ and all possible delays associated to the different presynaptic addresses are represented. 
Using this dense representation, the processing of $A$ by the model can be written as:

\begin{equation}\label{eq:kernel_b}
\forall x, y, \timev, \; \;
\mathcal{C}_\class(x, y, \timev)
= \sum_{p, \delta_x, \delta_y, \synapticdelay} \kernel_\class(p, \delta_x, \delta_y, \synapticdelay) \cdot A(p, x - \delta_x, y - \delta_y, \timev-\synapticdelay)
\end{equation}

% The computational brick used in equation~\eqref{eq:kernel_b} can be extended to every position, in space and time, of the boolean matrix $A$: equation~\eqref{eq:stream_b}
% \begin{equation}\label{eq:conv}
% \forall \presynaddr \in \presynaddrspace, \forall \timev \in T, \; \;
% \mathcal{C}_\class(\presynaddr,\timev)
% = \sum_{\ranksyn  \in [1, N_{\synapse^\postsynaddr}]} \kernel_\class(\presynaddr_\ranksyn, \synapticdelay_\ranksyn) \cdot A(\presynaddr, \timev-\synapticdelay_\ranksyn)
% \end{equation}
This shows that $\mathcal{C}_\class$ is a spatio-temporal convolution of the dense representation of the event stream with the dense kernels formed by the set of synapses:  $\mathcal{C}_\class = \kernel_\class \ast A$ (see Figure~\ref{fig:model} for an illustration). Note that to remain within the framework of a causal calculation, the kernels are shifted in time such that only past information gives an answer at the present time.
This well-known computation defines a differentiable measure which is very efficiently implemented for GPUs and which we will use for learning the classification of different motifs in the event stream. The same type of spatio-temporal filtering is used as a preprocessing stage of a pattern recognition algorithm~\citep{ghosh_spatiotemporal_2019}. Also,~\citet{sekikawa_constant_2018} developed an efficient 3D convolutional algorithm which implements a motion estimation task. By assuming a locally constant velocity, the authors assume the 3D kernel can be decomposed into a 2D kernel representing the shapes and a 3D kernel representing the velocity.

Here, we keep the analogy with spiking neurons, and we try to use as few heuristics as possible to observe the emergence of spiking motifs in the spatio-temporal kernels. If so, this precise spatio-temporal patterns prove to be of interest for neural computations and one can assume that biological neurons make use of this information as well. 
% \note{plus de biblio sur la motion detection/segmentation ou optique flow avec les events? (reviewer 2)}
The 3D convolution represents the linear integration of the spike train as input. The activation function of our model is a softmax function implementing a form of Multinomial Logistic Regression (MLR)~\citep{grimaldi_robust_2022}, in analogy to a spiking Winner-Take-All network~\citep{nessler_bayesian_2013}. In our MLR model, a probability value, for each class (i.e. each motion direction), is predicted for each event at position $x, y$ and at time $\timev$ as a softmax function of the linear combination of the list of events. % on the basal dendrite of a neuron $\postsynaddr$ in association to a specific class. The linear combination can be defined by a set of synapses $\synapse^\postsynaddr$ as described in the HD-SNN model. 
Using the kernels, it transforms the input raster plot into a probability with the following formula:

\begin{equation}\label{eq:mlr}
    \forall x, y, \timev, \forall \class \in [1, N_\class], \; \;
Pr(k=\class \; \vert \;  x, y, \timev) =
\frac {\exp  (\mathcal{C}_\class(x, y, \timev) +\bias_\class) }{Z(x, y, \timev)}
\end{equation}

where $Z(x, y, \timev)=\sum_{\class \in [1, N_\class]} \exp  (\mathcal{C}_\class(x, y, \timev) +\bias_\class)$ is the partition function and  $\bias_\class$ is the bias linked to class $\class$. 
In particular, we expect that some specific motifs may become tightly synchronized as they reach the basal dendritic tree, leading to a high postsynaptic activity which makes it progressively more likely to generate an output spike. The spiking output of the model corresponds to an event with the highest probability class.

Once explained the general framework used to solve the ecological task described in section~\ref{sec:task}, we can add some heuristics based on neuroscientific observations to constrain our model and its strategies to solve the task. Note that the general framework is the same as the one presented in~\cite{grimaldi_learning_2022} and that, appart from the task to solve and a deeper analysis of the results, these priors, inspired by neuroscience, are the main methodological differences with this previous study. We set \note{arbitrarily} the size of the kernels to $(9, 17, 17)$ and define as many classes as the number of motion direction: $N_\class = 8$. Because receptive fields of biological neurons are mainly circular and can be more or less elongated, we apply a circular mask on the spatial dimensions of the kernels. And because the synaptic delay is linked, by physical principles, to the distance between the presynaptic neuron and the postsynaptic one, we apply a linear mask on the temporal axis such that the connectivity depends on the delay: the lower the delay the lower the circular mask radius (see Figure~\ref{fig:kernels} for an illustration). The second heuristics is based on the observation that biological neurons have a sparse spiking activity. From this observation, we set an output firing rate for the layer of spiking neurons in order to make a sparse selection of the events at which the model makes a decision. It results in the selection of the coefficients with the maximum likelihood value. Therefore, we will only select a given proportion of the most active cells, which we set here to $.01\%$ of all output voxels \note{change accordingly to the model's parameters}. This threshold was found using cross-validation and future work should automate how this may be determined for various datasets. Given a stimulus input, the model thus yields a spike output in the postsynaptic address space.

@@

\begin{figure}%[h!]%[ht!]
    \centering
    \includegraphics[width=\linewidth]{figures/conv_HD-SNN.pdf}
    % \includegraphics[width=\linewidth]{figures/HD-SNN.png}
    \caption{
    \textbf{- HD-SNN illustration. From left to right:} A 3D representation of the event stream as output of the DVS. A 3D convolution is applied, on the dense representation of the input, along the $X$ and $Y$ spatial axes (defining the DVS pixel grid) and the temporal axis. There are as many kernels as the possible number of motion classes in the recording. For clarity, only $4$ kernels are illustrated and some weights are not represented in the voxel grid. For each voxel on the dense representation of the event stream as input, the output of the 3D convolution is processed by the non-linearity of the MLR model (i.e. softmax function). The output of the MLR gives a probability for each class, associated to a specific kernel, to emit an output event, or spike. With an appropriate spiking mechanism, here a winner-takes-all, we obtain, as output of the HD-SNN model, a new spike train with the different spikes associated to a specific motion class. }
    \label{fig:model}
\end{figure}

@@

Since the model is fully differentiable, we can now implement a semi-supervised learning rule. The loss function of the MLR model is the binary cross-entropy on the output of the classification layer. Supervision was implemented using the input binary events as defined above and the motion direction labels as the desired output. The labels were defined at each time point as a one-hot encoding of the current motion in the channel corresponding to the current motion for all positions. As in this semi-supervised context, the label is known, but the timing is not, we used a selection of the temporal support in an unsupervised manner. It is important to note that we weighted the cost function by considering only active cells, such that the error is only back propagated to the spatial locations of these most active cells. This is reminiscent of previous methods solving this problem using a Winner-Takes-All mechanism~\citep{masquelier_unsupervised_2007}. Simulations are performed with the \note{PyTorch library using gradient descent with Adam (for $2^{12}$ movies and a learning rate of $10^{-5}$)}. We demonstrated in a previous work that this type of learning algorithm can be assimilated to a Hebbian learning mechanism~\cite{grimaldi_robust_2022}.  
